---
title: "Exploratory analysis using group_by"
author: "Olya Besedina"
data: "10/03/2019"
output: github_document
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(viridis)

# copy the rest from github


```

# create data frame

```{r load data, cache=TRUE}
weather_df = 
  rnoaa::meteo_pull_monitors(c("USW00094728", "USC00519397", "USS0023B17S"),
                      var = c("PRCP", "TMIN", "TMAX"), 
                      date_min = "2017-01-01",
                      date_max = "2017-12-31") %>%
  mutate(
    name = recode(id, USW00094728 = "CentralPark_NY", 
                      USC00519397 = "Waikiki_HA",
                      USS0023B17S = "Waterhole_WA"),
    tmin = tmin / 10,
    tmax = tmax / 10,
    month = lubridate::floor_date(date, unit = "month")) %>%
  select(name, id, date, month, everything())
```

# group_by
```{r}
view(weather_df)
weather_df %>%
  group_by(name, month)
```


```{r}
weather_df %>% 
  group_by(name, month) %>% 
  summarize(n_obs = n())
```


# summirize by unique obs., n_distinct

```{r}
weather_df %>% 
  group_by(month) %>% 
  summarize(n_obs = n(),
            n_unique = n_distinct(date))
```

# count

```{r}
weather_df %>% 
  count(name)

```

# can rename
```{r}
weather_df %>%
  count(month, name = "n_days")
```

# dont use this. terrible

```{r, eval = FALSE}
weather_df %>% 
  pull(name) %>% 
  table()
```

# make a nice table for knited document

```{r}
weather_df %>% 
  count(name) %>% 
  knitr::kable()
```

## 2x2 table
a digression

Not the best way to do it
```{r}
weather_df %>% 
  filter(name != "Waikiki_HA") %>% 
  mutate(
    cold = case_when(
      tmax < 5 ~ "cold",
      tmax >= 5 ~ "not cold",
      TRUE    ~""
    )
  ) %>% 
  group_by(name, cold) %>% 
  count() %>% 
  pivot_wider (cold,
  values_from = n)
```

~ shows outcome

```{r}
weather_df %>% 
  filter(name != "Waikiki_HA") %>% 
  mutate(
    cold = case_when(
      tmax < 5 ~ "cold",
      tmax >= 5 ~ "not cold",
      TRUE    ~""
    )
  ) %>% 
  janitor::tabyl(name, cold)
```


Do logistic regression instead of 2x2 table

## general summaries 

```{r}
weather_df %>% 
  group_by(name, month) %>% 
  summarize(
    n = n(),
    mean_tmax = mean(tmax, na.rm = TRUE),
    sd_tmax = sd(tmax, na.rm = TRUE),
    median_prcp = median(prcp, na.rm = TRUE)
  )

```

na.rm = TRUE, if there is NA then drop it


```{r}
weather_df %>% 
  group_by(name, month) %>% 
  summarize(
    n = n(),
    mean_tmax = mean(tmax, na.rm = TRUE)) %>% 
  pivot_wider(
    names_from = name,
    values_from = mean_tmax
  ) %>% 
  knitr::kable(digits = 1)
```


  knitr::kable(digits = 1), only show one decimal in mean

# group mutates

```{r}
weather_df %>% 
  group_by(name) %>%
  ungroup()
```

# mutate per group
```{r}
weather_df %>% 
  group_by(name) %>% 
  mutate(
    mean_tmax = mean(tmax, na.rm = TRUE),
    center_tmax = tmax - mean_tmax) %>% 
  ggplot(aes(x = date, y = center_tmax, color = name))+
  geom_point()
```


```{r}
weather_df %>%
  group_by(name, month) %>%
  mutate(
    tmax_rank = min_rank(tmax)) %>% 
  filter(tmax_rank == 1) %>% 
  view()
  
```

lags and leads

lag function will show you value from previous day
```{r}
weather_df %>%
  group_by(name) %>%
  mutate(temp_change = tmax - lag(tmax)) %>%
  summarize(temp_change_sd = sd(temp_change, na.rm = TRUE),
            temp_change_max = max(temp_change, na.rm = TRUE))

```









